Ownership types in theory and practice (in Rust)

### 2023-01-15 paper/paper.tex
Finished memory safety example paragraph

@@

One such flavor is the \emph{owners-as-dominators model} \cite{ownership-types-survey}, which is an implementation of the original Flexible Alias Protection concept \cite{flexible-alias-protection}, which described a core encapsulation mechanism.

@@

Another kind is the owners-as-modifiers model \cite{ownership-types-survey}, which is a weaker form of owners-as-dominators that additionally allows read-only references.
A read-only reference can only be used to read files and call pure methods, which may not modify any existing objects.
In essence, aliasing is unrestricted, but only the owner of an object can modify it.
Owners-as-modifiers extend the rules specified for owners-as-dominators.

@@

% TODO: is it obvious that this contains exact sentences from citation?
The paper that first introduced the idea of owners-as-ombudsmen \cite{owners-as-ombudsmen} describes the concept as an extension of the owners-as-dominators model.
With owners-as-dominators, every path from the root to an object must pass through its owner.
Thus, encapsulated objects must have a single \emph{bridge object} that mediates all external interaction with its internal objects.
The owners-as-ombudsmen approach, on the other hand, relaxes the single bridge object constraint and allows several bridge objects to collectively define an aggregate with a shared representation.
These bridge objects are called \emph{ombudsmen} to emphasize their benevolent nature.

@@

\section{Implementation in Rust}
\label{sec:implementation-rust}

>>>>

\paragraph{Example}

Ralf Jung \cite{understanding-evolving-rust} provides a typical example of how ownership types can prevent a memory safety issue.

First, we will explore how a memory safety problem is created in the following C++ code snippet that does not use the ownership concept.

\begin{lstlisting}
std::vector<int> v { 10, 11 };
int *vptr = &v[1]; // points *into* v
v.push_back(12);
std::cout << *vptr; // bug (use-after-free)
\end{lstlisting}

We initialize a vector \emph{v} that contains two integers stored in a buffer in memory.
Next, we create a pointer \emph{vptr} that points into this buffer, specifically to the place where the second element (with current value 11) is stored.
Now, both \emph{v} and \emph{vptr} point to (overlapping parts of) the same buffer.
% TODO: change wording from source
We say that the two pointers are aliasing.
Then, we push a new element to the end of \emph{v}.
If the vector's capacity were large enough, the new element would be appended to the buffer.
However, we will assume there is no more space for an additional element, so a new buffer is allocated, and all the existing elements are moved over.
This case is interesting because \emph{vptr} still points to the old buffer!
In other words, adding a new element to \emph{v} has turned \emph{vptr} into a dangling pointer.
Therefore, trying to access the value the dangling pointer \emph{vptr} points to will cause a use-after-free bug.
% More generally, we can observe that an action through a pointer (\emph{v}) will also affect all of its aliases (\emph{vptr}).
The main problem here is that such memory issues cannot be detected at compile time.

% TODO: add graph from paper to visualize

Now, we will demonstrate how ownership types in Rust allow the compiler to detect such a use-after-free bug.
Consider the following Rust translation of our C++ example:

\begin{lstlisting}
let mut v = vec![10, 11];
let vptr = &mut v[1]; // points *into* v
v.push(12);
println!("{}", *vptr); // compiler error
\end{lstlisting}

Syntactically, the C++ and Rust versions are very similar.
Notably, the \emph{push} method in line 3 has the following signature: \verb|pub fn push(&mut self, value: T)|.
While the C++ version will compile the program successfully, causing a run-time security vulnerability, the Rust compiler shows an error: "Cannot borrow \emph{v} as mutable more than once at a time."
For now, it suffices to say that the Rust compiler disallows both lines 2 and 3 to create mutable references to \emph{v} if one of the mutable references is used later (like \emph{vptr} is in line 4).
In \ref{sec:rust-references}, we will explain in more detail how the Rust borrow-checker is able to detect this problem using Rust's ownership rules.

% A quick definition of ownership: you cannot have mutability and alias at the same time.
% My own example:

% \begin{lstlisting}
% fn dangle() -> &String {
%   let s = String::from("hello world");
%   &s
% }
% \end{lstlisting}

@@

One such flavor is the \emph{owners-as-dominators} model \cite{ownership-types-survey}, which is an implementation of the original Flexible Alias Protection concept \cite{flexible-alias-protection}, which described a core encapsulation mechanism.

@@

Another kind is the \emph{owners-as-modifiers} model \cite{ownership-types-survey}, which is a weaker form of owners-as-dominators that additionally allows read-only references.
A read-only reference can only be used to read files and call pure methods, which may not modify any existing objects.
In essence, aliasing is unrestricted, but only the owner of an object can modify it.
Owners-as-modifiers extend the rules specified for owners-as-dominators.

@@

% TODO: is it obvious that this contains exact sentences from citation?
The paper that first introduced the idea of \emph{owners-as-ombudsmen} \cite{owners-as-ombudsmen} describes the concept as an extension of the owners-as-dominators model.
With owners-as-dominators, every path from the root to an object must pass through its owner.
Thus, encapsulated objects must have a single \emph{bridge object} that mediates all external interaction with its internal objects.
The owners-as-ombudsmen approach, on the other hand, relaxes the single bridge object constraint and allows several bridge objects to collectively define an aggregate with a shared representation.
These bridge objects are called \emph{ombudsmen} to emphasize their benevolent nature.

@@

\section{Implementation in Rust}
\label{sec:implementation-in-rust}
